Investigating the Impact of Macroeconomic Indicators on the FTSE 250 Index Using Machine Learning Methods

### 2024-08-13 tex/paper.tex
intro, review, algorithm modified + references added

The stock market has played a central role in the economy and 
political ascent of many states ever since its inception. 
Accordingly, any instabilities that have arisen in the stock market 
throughout history have had far-reaching destabilising effects on the 
economy; the Mississippi Bubble (1717-1720) was an early example of the 
dangers that stock markets carry, where the French colonial administration 
in America went bankrupt as a result of a speculative bubble. Wth the
growing interconnectedness of markets, the stock market as an institution 
has also contributed significantly to global disasters such as the Great 
Depression and the Global Financial Crisis of 2008/9. Thus, economic 
policymakers and administrators must take into careful consideration 
both the impacts of their macroeconomic policies on the stock market.

Investors and traders are also affected by the 
fluctuations of economic factors due to their effect on firms.
The ability to forecast market trends allows 
investors to optimise their trading strategies, allowing them to make
informed decisions about when to buy or sell assets. 
Additionally, in a globalised economy, the interconnectedness of 
markets means that events in one region may have significant effects worldwide, 
making it even more critical for investors to stay ahead of potential market movements.
The traditional methods that have been leveraged in this context have been econometric, 
however machine learning methods have emerged as powerful alternatives capable
of capturing complex relationships in noisy financial data, and thereby providing more
accurate forecasts. Although econometric methods are still valuable in informing us
of the relevance of various macroeconomic variables.

We utilise the FTSE 250 as our measure of the UK stock market, ecause unlike the 
FTSE100 its overseas share of earnings is 
roughly $50\%$, similar to the S\&P 500. In contrast, nearly $80\%$ of the FTSE $100$'s earnings 
are from overseas. This is reflected in the constitution of the respective indices, with 
the FTSE 250 historically resembling the structure 
of the UK economy more closely than the FTSE100.

Thus, the study has a twofold aim:
\begin{enumerate}
    \item To investigate which macroeconomic variables influence the stock market price significantly, and quantify the effects of the following selection of relevant macroeconomic indicators on the FTSE 250, chosen as a more representative measure of the UK stock market. An Auto-regressive Distributed Lag (ARDL) model is utilised, in conjunction with an Error Correction Model (ARDL-ECM) over the January 1993 - March 2024 period: interest rate, inflation, USD exchange rate, and M3 money supply. 
    \item To investigate the efficacy of machine learning methods, in particular a radial basis function neural network, in predicting stock price movements based on the same macroeconomic indicators. 
\end{enumerate}

@@

The USD/GBP exchange rate has exhibited moderate variability, 
ranging from 0.4831 to 0.8834, suggesting periods of both 
appreciation and depreciation of the Sterling. The little variability 
there is can be
attributed to major events such as the 2008 financial crisis and Brexit, 
which both greatly affected the value of the Pound.

The Consumer Price Index (CPI), with an average of 88.36, has 
experienced significant inflationary fluctuations, ranging from 
62.8 to 133.5. These fluctuations highlight varying price levels and 
reflect periods of both high inflation, particularly in the late 2000s 
near the global financial crisis, and also during 2021-22 as the Russia-Ukraine
conflict destabilised energy and food markets, immediately after the devastation 
of supply chains by the COVID-19 pandemic; and also low inflation in the 
years of austerity in the UK and the global recession that onset during 
the COVID19 pandemic.

Interest rates, averaging $3.21\%$, have varied widely from $0.1\%$ to $8.4\%$, with a standard deviation of $2.52\%$, reflecting the 
increasing role of monetary policy in the UK. The Bank of England was forced to adjust interest rates 
many times in response to worsening economic conditions, examples being the steep rate cuts during the 2008 crisis to stimulate the economy and the gradual rate hikes in response to subsequent inflationary pressures.

The money supply (M3), averaging 1,861,958.47, has shown substantial variability, ranging from 494,181 to 3,704,873, 
in large part influenced by the various Quantitative Easing (QE) programs implemented by the Bank of England 
in response to the aforementioned crises. These programs aimed to inject liquidity into the economy, thus inflating the overall money supply.

@@

\begin{table}[h!]
    \centering
    \caption{Coefficients Estimates}
    \begin{tabular}{llll}
        \toprule
        \textbf{Variable} & \textbf{Long-Run Coefficient} & \textbf{Variable} & \textbf{Short-Run Coefficient} \\
        \midrule
        FTSE$_{250}$ & N/A & FTSE$_{250}$ $(-5)$  & $-0.033089$ \\
        CPI & $3.157179$ & CPI $(-4)$ & $-0.921235$ \\
        INT & $-0.1341097^{**}$ & INT $(-7)$ & $-0.034572^{****}$\\
        EXCHG &  $-0.890146^{***}$ & EXCHG $(-7)$ & $0.122318$ \\
        M3 & $-0.1141114$ & M3 $(-4)$ & $-0.527798^{***}$ \\
        Intercept & $-3.447767$ & ECT $(-1)$ & $-0.16^{**}$ \\
        \bottomrule
    \end{tabular}
\end{table}

@@

The ARDL Bounds Test confirmed that our variable selection was sensible by virtue
of its establishment of cointegration. The RBF-NN, it is hoped, will be able
to capture intricate non-linear patterns in the data that the econometric model
was unable to, and accordingly be able to predict with accuracy the future price
movements. The RBF-NN was utilised for both short-term and long-term 
predictions. For short-term forecasting, 
the model was tested over a 5-month horizon, based on the recommendation of the
ARDL model, while the long-term predictions spanned 57 months. 
We differenced the data based on the recommendations of the 
unit root tests conducted earlier to achieve stationarity.

>>>>

The stock market has played a central role in the economy and 
political ascent of many states ever since its inception. 
Accordingly, any instabilities that have arisen in the stock market 
throughout history have had far-reaching destabilising effects on the 
economy; the Mississippi Bubble (1717-1720) was an early example of the 
dangers that stock markets carry, where the French colonial administration 
in America went bankrupt as a result of a speculative bubble. With the
growing interconnectedness of markets, the stock market as an institution 
has also contributed significantly to global disasters such as the Great 
Depression and the Global Financial Crisis of 2008/9. Thus, economic 
policymakers and administrators require forecasts for how policy changes or 
unexpected economic trends may impact the economy through the stock market.

Investors and traders are also intimately affected by the 
fluctuations of economic factors due to their destabilising effect on firms.
The ability to forecast stock index price changes allows 
investors to optimise their trading strategies, enabling them to make
informed decisions about when to buy or sell assets. 
Additionally, in a globalised economy, the interconnectedness of 
markets means that events in one region may have significant effects worldwide, 
making it even more critical for investors to stay ahead of potential market movements.

The traditional methods that have been leveraged in both of these contexts have been econometric.
Machine learning methods, particularly neural networks, have emerged as powerful alternatives capable
of capturing complex relationships in financial data, and thereby providing more
accurate forecasts than econometric models.

In this study we first use an Autoregressive Distributed Lag (ARDL) model in conjunction with an Error Correction Model (ECM), to investigate 
the different effects of selected macroeconomic variables on the FTSE250 Index over the January 1993 - March 2024 period.
Second, we test the efficacy of a Radial Basis Function Neural Network (RBF-NN) in forecasting the price movements of the FTSE250 based on the selected macroeconomic indicators, over the short-term 
(5 months) and long-term (57 months).

We utilise the FTSE 250 as our measure of the UK stock market, because unlike the 
FTSE100 its overseas share of earnings is 
roughly $50\%$, similar to the S\&P 500. In contrast, nearly $80\%$ of the FTSE $100$'s earnings 
are from overseas. This is reflected in the constitution of the respective indices, with 
the FTSE 250 historically resembling the structure 
of the UK economy more closely than the FTSE100.

The macroeconomic variables we consider are inflation, indicated by the Consumer Price
Index (CPI); interest rate, represented by the Bank of England policy rate; the exchange rate, 
represented by the USD/GBP exchange rate; and the money supply, represented by the 
M3 measure of money supply, preferred over other 
measures of the money supply because it considers more assets than M2 money supply 
but does not include extremely illiquid assets as the M4 measure does. The selection 
of these variables was based on the literature and the structure of the UK economy since 
the establishment of the FTSE250 Index.

The remainder of the paper is organised as follows. 
Section 2 presents a review of the literature in this problem space, and explains the rationale behind our selection of macroeconomic variables.
Section 3 covers the data utilised in the study. Section 4 outlines the ARDL-ECM model, our methodology, and our results describing
the effect of the selected macroeconomic variables on the FTSE250. Section 5 constructs the RBF-NN model, and presents the short-term and long-term
forecasts for the FTSE250 Index price movements, based on our selected macroeconomic variables. Section 6 concludes the study.

@@

The USD/GBP exchange rate has exhibited moderate variability, exhibiting the stable relations between 
the two countries. Nevertheless, geopolitical shocks such as the 2008 financial crisis and Brexit both greatly affected the value of the Pound, 
which accounts for the large range.

The CPI, with an average of 88.36, has 
experienced significant fluctuations, highlight periods of high inflationary pressure such as in the late 2000s 
near the global financial crisis, during 2021-22 as the Russia-Ukraine, and periods of stable, low
inflation in the 2010s.

Interest rates, have varied widely from $0.1\%$ to $8.4\%$, with a standard deviation of $2.52\%$, reflecting the 
increasing role of monetary policy in the UK. The Bank of England was forced to adjust interest rates 
several times in response to worsening economic conditions, examples being the steep rate cuts during the 2008 financial crisis to stimulate the economy and the gradual rate hikes in response to 
subsequent post-COVID inflationary pressures.

The M3 money supply has shown substantial variability, ranging from 494,181 to 3,704,873, 
in large part influenced by the various Quantitative Easing (QE) programs implemented by the Bank of England 
in response to the aforementioned crises. These programs aimed to inject liquidity into the economy to stimulate economic activity, thus inflating the overall money supply.

@@

\begin{table}[h!]
    \centering
    \caption{Coefficients Estimates}
    \begin{tabular}{llll}
        \toprule
        \textbf{Variable} & \textbf{Long-Run Coefficient} & \textbf{Variable} & \textbf{Short-Run Coefficient} \\
        \midrule
        FTSE$_{250}$ & N/A & FTSE$_{250}$ $(-5)$  & $-0.033089$ \\
        CPI & $3.157179$ & CPI $(-4)$ & $-0.921235$ \\
        INT & $-0.1341097^{***}$ & INT $(-7)$ & $-0.034572^{****}$\\
        EXCHG &  $-0.890146^{****}$ & EXCHG $(-7)$ & $0.122318$ \\
        M3 & $-0.1141114$ & M3 $(-4)$ & $-0.527798^{***}$ \\
        Intercept & $-3.447767$ & ECT $(-1)$ & $-0.16^{**}$ \\
        \bottomrule
    \end{tabular}
\end{table}

@@

The RBF-NN was utilised for both short-term and long-term 
predictions. For short-term forecasting, 
the model was tested over a 5-month horizon, based on the optimum lag calculated earlier, while the long-term predictions spanned 57 months. 
We differenced the data based on the recommendations of the 
unit root tests conducted earlier to achieve stationarity.
